Higher order unification for free!

### 2024-04-06 paper/main.tex
some work on the intro

  The object logic we are interested in is Coq's~\cite{Coq-refman}
  Calculus of Inductive Constructions (CIC),
  for which we aim to implement a unification procedure \Uo using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory comprises
  $\eta\beta$ equivalence and comes equipped with a
  higher order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  We want \Uo to be as powerful as \Ue but on the object logic CiC.
  Elpi also comes with an encoding for CiC that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
  Unfortunately this encoding, which we refer to as \Fo,
  ``underuses'' \Ue by restricting it to first-order unification problems only.
  To address this issue, we propose a better-behaved encoding, \Ho,
  demonstrate how to map unification problems in \Fo{}
  to related problems in \Ho, and illustrate
  how to map back the unifiers found by \Ue, effectively implementing
   \Uo on top of \Ue for the encoding \Fo.

@@

Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
which have been utilized to implement various formal systems such as
First Order Logic~\cite{felty88cade},
Set Theory~\cite{10.1007/BF00881873},
Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
Constructions~\cite{felty93lics}.
The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of inductive Constructions (Cic), and we want to code a
type-class~\cite{wadler89} solver for Coq~\cite{Coq-refman} using
the Coq-Elpi~\cite{tassi:hal-01637063}
meta programming framework.
Type-class solvers are unification based proof search procedures
that combine a set of designated lemmas in order to providing essential
automation to widely used Coq libraries.

As an example we take the \coqIn{Decide} type class
from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
library. The class identifies predicates equipped with a decision procedure.
The following three designated lemmas (called \coqIn{Instances} in the
type-class jargon) state that: 1) the type \coqIn{fin n}, of natural numbers
smaller than \coqIn{n}, is finite; 2) the predicate \coqIn{nfact n nf},
linking a natural number \coqIn{n} to its prime factors \coqIn{nf}, is decidable;
3) the universal closure of a predicate has a decision procedure if
its domain is finite and if the predicate is decidable.

@@

\noindent Under this context of instances a type-class solver is able to prove
the following statement automatically by back-chaining the lemmas above:

\begin{coqcode}
  Check _ : Decision (forall y: fin 7, nfact y 3).       ~\customlabel{goal:g}{(g)}~
\end{coqcode}

\noindent
The encoding of CiC provided by Elpi, that we will discuss at length later in
\cref{sec:encodings,sec:lang-spec}, is an Higher Order Abstract
Syntax (HOAS) datatype \elpiIn{tm} featuring (among others) the following
constructors:

@@

\noindent
we obtain an unsolvable unification problem \ref{problem:p}:
the two lists of terms have different lengths!
%The root cause is that
%\ref{problem:p} is an higher order in CiC, but becomes
%first order in the meta language due to the ``naive'' encoding.

@@

To state precisely the problem we solve we need a \Fo{} representation
of CiC terms and a \Ho one.
We call \Eo the equality over ground terms in \Fo,
\Ee the equality over ground terms in \Ho,
\Uo the unification procedure we want to implement and
\Ue the one provided by the meta language.
\noteE[inline]{extend \Eo and \Ee with reflexivity on uvars.}

@@

Our encoding of CiC may look ``semi shallow'' since we use the meta-language
lambda abstraction but not its application (for the terms of type \elpiIn{tm}).
A fully shallow encoding unfortunately does not fit our use case, although
it would make the running example work:

@@

\noindent
There are two reasons for dismissing this encoding. The first one is that
in CiC it is not always possible to adopt it since the type system
of the meta language is too weak to accommodate terms with a variable arity,
like the following example:

@@

In the literature we could find a few related encoding of CiC.
TODO In~\cite{felty93lics} is related and make the
discrepancy between the types of ML and CiC visible. In this case
one needs 4 application nodes. Moreover the objective is an encoding
of terms, proofs, not proof search. Also note the conv predicate,
akin to the unif we rule out.

@@

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first class terms, i.e. they have a concrete syntax. We keep these languages
minimal, for example, we omit the \elpiIn{all} quantifier of CiC we used
in the example in Section~\ref{sec:intro} together with the type notation of
terms carried by the \elpiIn{lam} constructor.
%
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}

@@

\section{Actual implementation}

In this paper we show a minimized example. The full code is there.
But we also have to code things in Coq.
In Elpi we don't have a main loop, we rely on the interpreter one.
So links are constraints and progress are CHR rules.

\begin{elpicode}
link-eta bla :- suspend-condition, !, declare_constraint (link-eta bla) Vars.
link-eta bla :- progress. % solves the goal, hence the constraint is not more in L.
constraints link-beta {
  % undup
  rule (G1 ?- link-eta (uvar X LX1) T1) /
       (G2 ?- link-eta (uvar X LX2) T2) |
       relocate LX1 LX2 T2 T2' <=>
       (G1 ?- T1 = T2').
}
\end{elpicode}

>>>>

  The object logic we are interested in is Coq's~\cite{Coq-refman}
  Calculus of Inductive Constructions (CIC),
  for which we aim to implement a unification procedure \Uo using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory comprises
  $\eta\beta$ equivalence and comes equipped with a
  higher order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  We want \Uo to be as powerful as \Ue but on the object logic CIC.
  Elpi also comes with an encoding for CIC that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
  Unfortunately this encoding, which we refer to as \Fo,
  ``underuses'' \Ue by restricting it to first-order unification problems only.
  To address this issue, we propose a better-behaved encoding, \Ho,
  demonstrate how to map unification problems in \Fo{}
  to related problems in \Ho, and illustrate
  how to map back the unifiers found by \Ue, effectively implementing
   \Uo on top of \Ue for the encoding \Fo.

@@

Meta languages such as Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
have been utilized to specify various
logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
In some cases, the most notable one being Higher Order Logic~\cite{books/sp/NipkowPW02},
the ML Isabelle is such a good fit that it implements an interactive proof
system for HOL, and not just a specification.

The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of Inductive Constructions (CIC), and we want to implement a
type-class~\cite{wadler89} solver for Coq~\cite{Coq-refman} using
the Coq-Elpi~\cite{tassi:hal-01637063} meta programming language,
a dialect of $\lambda$Prolog already used to extend
Coq~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
Type-class solvers are unification based proof search procedures
that  provide essential automation to widely used Coq libraries.
These solvers are reminiscent of Prolog: they back-chain lemmas taken
from a designated database of type class instances.

As an example we take the \coqIn{Decision} type class
from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
library. The class
identifies predicates equipped with a decision procedure.
The following three designated \coqIn{Instances} state that:
1) the type \coqIn{fin n}, of natural numbers
smaller than \coqIn{n}, is \coqIn{Finite} (another type class);
2) the predicate \coqIn{nfact n nf}, linking a natural number
\coqIn{n} to its prime factors \coqIn{nf}, is decidable;
3) the universal closure of a predicate has a decision procedure if
its domain is finite and if the predicate is decidable.

@@

\noindent Given this database of instances a type-class solver is
able to prove the following statement automatically,
by back-chaining the lemmas above:

\begin{coqcode}
  Check _ : Decision (forall n: fin 7, nfact n 3).       ~\customlabel{goal:g}{(g)}~
\end{coqcode}

\noindent
The encoding of CIC provided by Elpi, that we will discuss at length later in
\cref{sec:encodings,sec:lang-spec}, is an Higher Order Abstract
Syntax (HOAS) datatype \elpiIn{tm} featuring (among others) the following
constructors:

@@

\noindent
we obtain an unsolvable unification problem \ref{problem:p}:
the two lists of terms have different lengths!
%The root cause is that
%\ref{problem:p} is an higher order in CIC, but becomes
%first order in the meta language due to the ``naive'' encoding.

@@

To state precisely the problem we solve we need a \Fo{} representation
of CIC terms and a \Ho one.
We call \Eo the equality over ground terms in \Fo,
\Ee the equality over ground terms in \Ho,
\Uo the unification procedure we want to implement and
\Ue the one provided by the meta language.
\noteE[inline]{extend \Eo and \Ee with reflexivity on uvars.}

@@

Our encoding of CIC may look ``semi shallow'' since we use the meta-language
lambda abstraction but not its application (for the terms of type \elpiIn{tm}).
A fully shallow encoding unfortunately does not fit our use case, although
it would make the running example work:

@@

\noindent
There are two reasons for dismissing this encoding. The first one is that
in CIC it is not always possible to adopt it since the type system
of the meta language is too weak to accommodate terms with a variable arity,
like the following example:

@@

In the literature we could find a few related encoding of CIC.
TODO In~\cite{felty93lics} is related and make the
discrepancy between the types of ML and CIC visible. In this case
one needs 4 application nodes. Moreover the objective is an encoding
of terms, proofs, not proof search. Also note the conv predicate,
akin to the unif we rule out.

@@

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first class terms, i.e. they have a concrete syntax. We keep these languages
minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
in the example in Section~\ref{sec:intro} together with the type notation of
terms carried by the \elpiIn{lam} constructor.
%
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}

@@

\section{Actual implementation in Elpi}

In this paper we show a minimized example. The full code is there.
But we also have to code things in Coq-Elpi.

In Elpi we don't have a main loop, we rely on the interpreter one.
So links are constraints and progress are CHR rules.
Constraints are suspended goals that are resumend when some unif variable
is assigned.

\begin{elpicode}
link-eta bla :- suspend-condition Holes, !, declare_constraint (link-eta bla) Holes.
link-eta bla :- progress. % solves the goal, hence the constraint is not more in L.
\end{elpicode}

This matches with all progress rules but for the ones considering
two links. For these we need to resort to CHR to manipulate L. For example
deduplicate eta

\begin{elpicode}
constraints link-eta {
  rule (N1 : G1 ?- link-eta (uvar X LX1) T1) % match
    /  (N2 : G2 ?- link-eta (uvar X LX2) T2) % remove
    |  (relocate LX1 LX2 T2 T2')             % condition
   <=> (N1 : G1 ?- T1 = T2').                % new goal
}
\end{elpicode}

Remark how the invariant about uvar arity makes this easy, since LX1 and LX2
have the same length. Also note that N1 only contains the names of the first
link (while relocate runs in the disjoint union) and Elpi ensures that
T2' can live in N1.
